

# Results

```{r setup3, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
 
 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
```

```{r}
#| label: source1
#| echo: false
#| results: false
#| message: false
#| warning: false
#| include: false

library(here)
source(here("processing/prod_analysis_multinom.R"))
```


## Descriptive analysis


@fig-alluvial shows the annual frequencies of preferences for pension commodification in 2016, 2018, and 2023. Each year presents stacked percentages frequencies for the level of agreement and disagreement. Overall, a large majority rejects market-based access to old-age pension benefits, but this opposition has eased over time: 83.5% in 2016, 81.2% in 2018, and 71.8% in 2023. The 2016–2018 change is modest, whereas 2023 registers a 9.4 point drop in disagreement relative to 2018, mirrored by rising agreement: 16.5% (2016), 18.8% (2018), 28.2% (2023). Substantively, while most respondents continue to oppose the idea that higher-income individuals should obtain better pensions via the market, a non-trivial and growing portion endorses this statement, with the sharpest expansion concentrated in the latest wave (+9.4 points from 2018 to 2023). 


```{r}
#| label: fig-alluvial
#| fig-cap: "Change in preferences for pension commodification over time (2016, 2018, and 2023)."
#| fig-cap-location: bottom
#| echo: false
#| results: asis
g1 +
  theme(legend.text = element_text(size = 10))
```

Regarding the relationship between preferences for the commodification of pensions and intergenerational occupational mobility, @fig-mean shows the percentage of agreement with market-based access to pension benefits according to mobility trajectories and survey waves. Averaging across waves, agreement is highest among the High-High immobile (27.8%), followed by the Middle-High upwardly mobile (24.6%) and the High-Middle downwardly mobile (24.0%). The lowest levels are found among ascending Low-High (17.3%), and immobile Low-Low (17.4%) groups. The specific profiles of each wave accentuate these gradients in most trajectories, especially in 2023: agreement reaches 31.9% for High-High, 29.4% for upward Middle-High, and peaks at 36.8% for downward High-Middle (the highest of all trajectories for that year). Descriptively, immobility at the top is associated with greater support for commodification; among those who move, support is greatest for the upward Middle-High trajectory and the downward High-Middle trajectory, patterns that intensify in 2023. These shifts anticipate the heterogeneity documented below.


```{r}
#| label: fig-mean
#| fig-cap: "Percentage agreement on preferences for pension commodification according to mobility trajectory and time (2016, 2018, and 2023)."
#| out-width: '90%'
#| fig-asp: 0.7
#| fig-cap-location: bottom
#| echo: false
#| results: asis

g2 +
  theme(legend.text = element_text(size = 10))

```


## Mobility effects models

@fig-reg reports the double-robust estimates of the effect of intergenerational occupational mobility on preferences for pension commodification, obtained from weighted models that combine stabilized IPW with covariate adjustment and fixed wave effects. Results reveal a clear directional asymmetry between upward and downward mobility trajectories. Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]). Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high socioeconomic status—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle socioeconomic status. Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level. In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-status origins express weaker preferences for market-based pension benefits. The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts. These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them. 


```{r}
#| label: fig-reg
#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification. Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility. Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual; bars show 95% confidence intervals."
#| out-width: '90%'
#| fig-asp: 0.7
#| fig-cap-location: bottom
#| echo: false
#| results: asis

g3
```



## The role of meritocracy


To probe whether meritocratic beliefs channel the impact of mobility on preferences for pension commodification, I estimate interaction models between mobility and meritocratic beliefs and examine the conditional treatment effects of mobility at low and high levels of meritocracy (see [Supplementary Material](#anexo) for complete models).


```{r}
#| label: fig-interact
#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by meritocratic perception. Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility, estimated from interaction models (T × Merit). Points show the marginal effects (simple slopes) of the mobility treatment at low and high merit; bars are 95% confidence intervals. Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual."
#| out-width: '90%'
#| fig-asp: 0.8
#| fig-cap-location: bottom
#| echo: false
#| results: asis


g4+
  theme(legend.text = element_text(size = 10))
```

@fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic beliefs. Across trajectories, mobility effects among low-merit respondents are uniformly small, imprecise, and centered around zero for both upward and downward moves. Among high-merit respondents, only upward mobility from middle to high status shows a clear pro-commodification pattern, with a positive and statistically significant effect on MJP for pensions ($\beta = 0.22$, 95% CI [0.05, 0.38]). All other trajectories—Low→Middle, Low→High, Middle→Low, High→Middle, and High→Low—remain statistically indistinguishable from zero. Nonetheless, the point estimates broadly align with the theoretical expectations for upward moves: upwardly mobile individuals with stronger meritocratic beliefs tend to display higher support for pension commodification than their low-merit counterparts. By contrast, downward trajectories exhibit a pattern opposite to the mechanism hypothesis, with high-merit respondents generally showing more de-commodifying attitudes than those with weaker meritocratic beliefs. However, formal interaction tests do not detect robust heterogeneity in mobility effects by meritocratic beliefs (see [Supplementary Material](#anexo) for full interaction models): even for the Middle→High trajectory, the simple effect under high meritocracy does not differ significantly from its low-merit counterpart. Taken together, these results do not provide convincing support for the mechanism hypothesis: meritocratic beliefs do not operate as a consistent or general channel through which intergenerational mobility shapes preferences for the commodification of pensions.



## Robustness check and sensitivity analysis

As a robustness check, I re-estimate the models under three alternative codings of the outcome: (i) recoding the neutral category as agreement (3–5 vs. 1–2), (ii) excluding neutral responses from the analysis, and (iii) using the original 5-point item in a linear specification. Across these alternative models (see [Supplementary Material](#anexo) for complete tables), the substantive pattern of the main effects is preserved: the Middle→High trajectory remains positively associated with support for pension commodification, and the High→Low trajectory remains negatively associated with it, with both coefficients generally increasing in magnitude. By contrast, the negative High→Middle effect becomes statistically indistinguishable from zero in the robustness specifications, indicating that this particular estimate is more sensitive to how the outcome is coded than the other two trajectories.


To assess the extent to which the estimated mobility effects may be driven by unmeasured confounding, I implement the sensitivity analysis proposed by Cinelli and Hazlett [-@cinelli_making_2020], which quantifies how strong an unobserved confounder would need to be—in terms of its joint explanatory power for both treatment and outcome—to attenuate or explain away the ATT, using gender as a benchmark covariate (the predictor with the largest partial association with the outcome aside from treatment). For the Middle→High trajectory ($\beta \approx 0.09$), partial $R^2 \approx 0.013$, an unobserved confounder would need to explain about 10–11% of the residual variance of both treatment and outcome to reduce the effect to zero and roughly 4% to render it statistically insignificant, substantially more than gender does. For the High→Low effect ($\beta \approx -0.10$), partial $R^2 \approx 0.012$, the required strength is of similar magnitude (around 10–11% to explain away the effect and about 4% to remove significance), again exceeding the explanatory power of the main observed covariates. By contrast, the High→Middle effect ($\beta \approx -0.07$), partial $R^2 \approx 0.007$ is somewhat less robust: a confounder explaining around 8% of residual variance in both treatment and outcome could nullify the estimate, and one explaining about 2% could make it non-significant. Overall, the sensitivity analysis suggests that the positive Middle→High and negative High→Low effects are moderately robust to unmeasured confounding, whereas the High→Middle effect is more vulnerable to relatively modest omitted variables (see [Supplementary Material](#anexo) for complete tables).

