

# Results

```{r setup3, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
 
 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
```

```{r}
#| label: source1
#| echo: false
#| results: false
#| message: false
#| warning: false
#| include: false

library(here)
source(here("processing/prod_analysis_multinom.R"))
```


## Descriptive statistics


@fig-alluvial shows the annual frequencies of preferences for pension commodification in 2016, 2018, and 2023. Each year presents stacked percentages frequencies for the level of agreement and disagreement. Overall, a large majority rejects market-based access to old-age pension benefits, but this opposition has eased over time: 83.5% in 2016, 81.2% in 2018, and 71.8% in 2023. The 2016–2018 change is modest, whereas 2023 registers a 9.4 point drop in disagreement relative to 2018, mirrored by rising agreement: 16.5% (2016), 18.8% (2018), 28.2% (2023). Substantively, while most respondents continue to oppose the idea that higher-income individuals should obtain better pensions via the market, a non-trivial and growing portion endorses this statement, with the sharpest expansion concentrated in the latest wave (+9.4 points from 2018 to 2023). 


```{r}
#| label: fig-alluvial
#| fig-cap: "Change in preferences for pension commodification over time (2016, 2018, and 2023)."
#| fig-cap-location: bottom
#| echo: false
#| results: asis
g1 +
  theme(legend.text = element_text(size = 10))
```

Regarding the relationship between preferences for the commodification of pensions and intergenerational occupational mobility, @fig-mean shows the percentage of agreement with market-based access to pension benefits according to mobility trajectories and survey waves. Averaging across waves, agreement is highest among the High-High immobile (27.8%), followed by the Middle-High upwardly mobile (24.6%) and the High-Middle downwardly mobile (24.0%). The lowest levels are found among ascending Low-High (17.3%), and immobile Low-Low (17.4%) groups. The specific profiles of each wave accentuate these gradients in most trajectories, especially in 2023: agreement reaches 31.9% for High-High, 29.4% for upward Middle-High, and peaks at 36.8% for downward High-Middle (the highest of all trajectories for that year). Descriptively, immobility at the top is associated with greater support for commodification; among those who move, support is greatest for the upward Middle-High trajectory and the downward High-Middle trajectory, patterns that intensify in 2023. These shifts anticipate the heterogeneity documented below.


```{r}
#| label: fig-mean
#| fig-cap: "Percentage agreement on preferences for pension commodification according to mobility trajectory and time (2016, 2018, and 2023)."
#| out-width: '90%'
#| fig-asp: 0.7
#| fig-cap-location: bottom
#| echo: false
#| results: asis

g2 +
  theme(legend.text = element_text(size = 10))

```


## Mobility effects models

@fig-reg reports the double-robust estimates of the effect of intergenerational occupational mobility on preferences for pension commodification, obtained from weighted models that combine stabilized IPW with covariate adjustment and fixed wave effects. Results reveal a clear directional asymmetry between upward and downward mobility trajectories. Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]). Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high socioeconomic status—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle socioeconomic status. Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level. In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-status origins express weaker preferences for market-based pension benefits. The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts. These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them. 


```{r}
#| label: fig-reg
#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification. Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility. Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual; bars show 95% confidence intervals."
#| out-width: '90%'
#| fig-asp: 0.7
#| fig-cap-location: bottom
#| echo: false
#| results: asis

g3
```



## Effect heterogeneity


To probe potential mechanisms underlying the mobility effects, I estimate origin-specific ATT models that interact the mobility treatment with high meritocratic perceptions (ref. = low), using stabilized IPW weights, covariate adjustment, and wave fixed effects. I also estimate an analogous specification interacting mobility with survey waves (ref. = 2016) to assess temporal heterogeneity in the treatment effect (see [Supplementary Material](#anexo) for complete models).


```{r}
#| label: fig-interact
#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by meritocratic perception. Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility, estimated from interaction models (T × Merit). Points show simple slopes of the mobility effect at low merit and high merit; bars are 95% confidence intervals. Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual."
#| out-width: '90%'
#| fig-asp: 0.8
#| fig-cap-location: bottom
#| echo: false
#| results: asis

g4+
  theme(legend.text = element_text(size = 10))
```

@fig-interact shows that neither the interaction terms nor the mobility coefficients reach conventional significance across trajectories, indicating no detectable moderation of the causal mobility effect by meritocratic perceptions at the 95% level. Directionally, in the left panel, the interaction is positive for Middle→High ($\beta$ = 0.15, 95% CI [−0.03, 0.33]) and Low→High ($\beta$ = 0.05, [−0.14, 0.24]), suggesting, if anything, somewhat larger pro-commodification effects among high-merit respondents when upwardly mobile, whereas it is negative for High→Low ($\beta$ = −0.14, [−0.38, 0.10]) and High→Middle ($\beta$ = −0.03, [−0.23, 0.17]), showing weaker pro-commodification effects under downward mobility at high merit. Focusing on the treatment effect at low meritocracy (the reference category) at the right panel, the mobility coefficient captures the origin-specific ATT comparing movers to their immobility counterfactual among low-merit respondents. These effects are small and imprecise across trajectories: Low→Middle ($\beta$ = −0.02, [−0.09, 0.04]) and Low→High ($\beta$ = −0.03, [−0.10, 0.05]) are near zero. Middle→High is positive but not significant ($\beta$ = 0.07, [−0.01, 0.14]), Middle→Low is null ($\beta$ = -0.00, [−0.07, 0.06]), and downward from high shows negative, borderline estimates: High→Middle ($\beta$ = −0.07, [−0.15, 0.00]) and High→Low ($\beta$ = −0.08, [−0.15, 0.00]). Substantively, I do not find evidence that meritocratic perceptions moderate the causal effect of mobility, although the signs align with the idea that upward mobility at high merit tilts attitudes toward commodification, and downward mobility at high merit tilts them away.


```{r}
#| label: tbl-inttime
#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by survey wave"
#| tbl-cap-location: top
#| results: asis
#| echo: false

ccoef <- list(
  "(Intercept)" = "Intercept",
  "t" = "Mobility treatment",
  "ola2018" = "Wave 2018",
  "ola2023" = "Wave 2023",
  "t:ola2018" = "Mobility treatment x Wave 2018",
  "t:ola2023" = "Mobility treatment x Wave 2023"
  )

texreg::texreg(list(mp_lm_t,
               mp_lh_t,
               mp_mh_t,               
               mp_ml_t,
               mp_hm_t,
               mp_hl_t),
          custom.model.names = c(
            "Low-Middle",
            "Low-High",
            "Middle-High",            
            "Middle-Low",
            "High-Middle",
            "High-Low"
          ),   caption.above = NULL,
               caption = NULL,
               custom.coef.map = ccoef,
               digits = 2,
               groups = list("Wave (Ref.= 2016)" = 3:4, "Mobility treatment x Wave (Ref.= 2016)" = 5:6),
               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.80,
               center = T,float.pos = "H!",
               custom.gof.rows = list("Controls"=rep("Yes",6)))
               
```


@tbl-inttime shows the causal heterogeneity of the mobility effect across the different waves of the survey. Regarding temporal heterogeneity, the only statistically significant interaction term is observed for the High→Middle trajectory in 2023 ($\beta_{\text{int}} = 0.19$), 95% CI [0.01, 0.38]). However, when this interaction is added to the baseline effect in 2016 (-0.19), the marginal effect of High→Middle mobility in 2023 is approximately zero, and its confidence interval includes the null. Thus, rather than indicating a reversal toward a positive effect, the evidence suggests that the negative impact of downward mobility from high backgrounds is confined to 2016 and attenuates over time. No comparable temporal changes are observed in the remaining trajectories. Overall, the results provide little support for strong time-conditional heterogeneity in mobility effects, beyond the fading of the initially negative High→Middle effect.


## Robustness check and sensitivity analysis

As a robustness check, I re-estimate the models under three alternative codings of the outcome: (i) recoding the neutral category as agreement (3–5 vs. 1–2), (ii) excluding neutral responses from the analysis, and (iii) using the original 5-point item in a linear specification. Across these alternative models (see [Supplementary Material](#anexo) for complete tables), the substantive pattern of the main effects is preserved: the Middle→High trajectory remains positively associated with support for pension commodification, and the High→Low trajectory remains negatively associated with it, with both coefficients generally increasing in magnitude. By contrast, the negative High→Middle effect becomes statistically indistinguishable from zero in the robustness specifications, indicating that this particular estimate is more sensitive to how the outcome is coded than the other two trajectories.


To assess the extent to which the estimated mobility effects may be driven by unmeasured confounding, I implement the sensitivity analysis proposed by Cinelli and Hazlett [-@cinelli_making_2020], which quantifies how strong an unobserved confounder would need to be—in terms of its joint explanatory power for both treatment and outcome—to attenuate or explain away the ATT, using gender as a benchmark covariate (the predictor with the largest partial association with the outcome aside from treatment). For the Middle→High trajectory ($\beta \approx 0.09$), partial $R^2 \approx 0.013$, an unobserved confounder would need to explain about 10–11% of the residual variance of both treatment and outcome to reduce the effect to zero and roughly 4% to render it statistically insignificant, substantially more than gender does. For the High→Low effect ($\beta \approx -0.10$), partial $R^2 \approx 0.012$, the required strength is of similar magnitude (around 10–11% to explain away the effect and about 4% to remove significance), again exceeding the explanatory power of the main observed covariates. By contrast, the High→Middle effect ($\beta \approx -0.07$), partial $R^2 \approx 0.007$ is somewhat less robust: a confounder explaining around 8% of residual variance in both treatment and outcome could nullify the estimate, and one explaining about 2% could make it non-significant. Overall, the sensitivity analysis suggests that the positive Middle→High and negative High→Low effects are moderately robust to unmeasured confounding, whereas the High→Middle effect is more vulnerable to relatively modest omitted variables (see [Supplementary Material](#anexo) for complete tables).

